Spatial variation of the native colon microbiota in healthy adults

### 2017-11-02 submission/manuscript.tex
made all text and most figure edits

2. Pennsylvania State University, Hershey, Pennsylvania ??

@@

@@

\textbf{Background and Aims:} The microbiome has been implicated in the
development of colorectal cancer (CRC) and inflammatory bowel diseases
(IBD). The specific traits of these diseases vary along the axis of the
digestive tract. Further, variation in the structure of the gut
microbiota has been associated with both diseases. Here we profiled the
microbiota of the healthy proximal and distal mucosa and lumen to better
understand how bacterial populations vary along the colon.

\textbf{Methods:} We used a two-colonoscope approach to sample proximal
and distal mucosal and luminal contents from the colons of 20 healthy
subjects that had not undergone any bowel preparation procedure. The
biopsies and home-collected stool were subjected to 16S rRNA gene
sequencing and Random Forest classification models were built using taxa
abundance and location to identify microbiota specific to each site.

\textbf{Results:} : The right mucosa and lumen had the most similar
community structures of the five sites we considered from each subject.
The distal mucosa had higher relative abundance of Finegoldia,
Murdochiella, Peptoniphilus, Porphyromonas and Anaerococcus. The
proximal mucosa had more of the genera Enterobacteriaceae, Bacteroides
and Pseudomonas. The classification model performed well when
classifying mucosal samples into proximal or distal sides (AUC=0.850).
Separating proximal and distal luminal samples proved more challenging
(AUC=0.580) and specific microbiota that differentiated the two were
hard to identify.

\textbf{Conclusions:} By sampling the unprepped colon, we identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to different disease processes on their respective
sides of the colon.

@@

\textbf{Keywords}: microbiome; colon cancer; proximal and distal colon

@@

Luminal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Figure 1). Subjects also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on DNA extracted from each
sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Figure 2A) (14). Samples had
varying levels of diversity at each site, irrespective of the individual
(Figure 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

To compare similarity between the proximal and distal sides and within
the lumen and mucosa, we compared the community structure of these sites
based on the relative abundances of individual Operational Taxonomic
Units (OTUs). Across all subjects we observed wide variation when
comparing sample locations (Figure 3A). Those ranges did not follow a
clear pattern on an individual basis. However, when comparing median
dissimilarity between the communities found in the proximal lumen and
mucosa, the proximal and distal lumen, the proximal and distal mucosa,
and the distal lumen and mucosa, we found that the proximal lumen and
mucosa were most similar to each other than to the other samples (P
\textless{} 0.005, Wilcoxon, BH adjustment).

Next, we compared the luminal and mucosal samples to the fecal sample of
each subject. Amidst the large inter-subject variation, we did identify
significantly less dissimilarity between the distal luminal sample and
the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
Furthermore, there was an even larger difference in the communities
found in the distal mucosa compared to the fecal communities, indicating
that the mucosa is as different from the stool as compared to lumen (P
\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
the contents of the distal lumen were most representative of the
subjects' feces, and the mucosal microbiota are distinct from the fecal
and luminal communities.

To determine what factors may have driven the differences seen among the
samples, we compared the community dissimilarity between samples from
all subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to matched samples from the other subjects
(Figure 3C); this is consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (15--17). Thus
interpersonal variation drove the differences between samples more than
whether the sample came from the proximal or distal side of the colon or
from the lumen or mucosa.

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We used
10-fold cross validation to build the first model to classify the
luminal versus mucosal samples for the proximal and distal sides,
independently (Figure S1A). The models performed well when classifying
these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
were most predictive of each site were identified by their greatest mean
decrease in accuracy when removed from the model. For distinguishing the
proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
included in the best model (Figure 4A). The model to differentiate
between the distal lumen and mucosa included OTUs affiliated with the
\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
\emph{Anaerococcus} (Figure 4B). These results indicated that there were
fine differences between the different sites of the colon, and that
these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples using 10-fold cross validation. The model
performed best when distinguishing the proximal versus distal mucosa
(Figure S1B, AUC = 0.850) whereas the model to differentiate between the
proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
included in the model differentiating the distal and proximal mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Figure
5A). The model that attempted to separate the the proximal and distal
lumen included OTUs affiliated with the \emph{Bacteroides},
\emph{Clostridium IV} and \emph{Oscillibacter} (Figure 5B).
Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were distinct
between the mucosa and lumen and also helped to differentiate between
the proximal and distal sides.

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors
(18). There were 14 samples from 8 subjects with the \emph{F. nucleatum
subsp. animalis} sequences. Of the samples with the highest relative
abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
were from the proximal mucosa and three from the distal mucosa (Figure
S2A). The second most frequent \emph{Fusobacterium} sequence was OTU472,
which aligned with 99\% identity to \emph{F. varium}. In addition to
\emph{F. nucleatum}, \emph{F. varium} has been associated with IBD (19).
Four subjects harbored \emph{F. varium} and the samples were split
evenly between the proximal and distal mucosa (Figure S2B). OTU152 was
similar to the members of the \emph{Porphyromonas} genus and the most
frequent sequence in that OTU aligned to \emph{Porphyromonas
asacharolytica} (99\% over full length), another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the distal mucosa, and in fact was one of the OTUs the classification
model identified as separating distal and proximal sides (Figure S2C).
Among the 11 distal mucosa samples that were positive for \emph{P.
asacharolytica}, the relative abundances for this OTU ranged from 0.01\%
to 16\%. Thus, disease-associated OTUs could be found in our sample set
of 20 healthy individuals.

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, casued by differences in age, sex, diet, but also
spatial ``patchiness'' in the gut microbiome. One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(20). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (21). That
said, across our samples, the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to luminal content in humans (4). Hence, the conclusions
we were able to draw from phyla analysis may have been impacted by
inter-subject patchiness.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (22). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} along the
colon, but given its anaerobic lifestyle and asacharolytic metabolism,
it is not surprising that it resides in the less-oygen-rich and
protein-rich distal mucosa (4). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreased along the colon to rectum (23). In another study,
\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
mutations, molecular features of proximal CRC (24). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in the present study, the
bacterium was spread across the proximal mucosa, distal lumen and distal
mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
suggests that \emph{Fusobacteria} species are more commonly found on
proximal tumors and in biofilms, indicating that it is not only the
presence of the bacterium but the structure of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
only co-occur on CRC tumors but also to synergize to promote
tumorigenesis in an oral cancer model (25) (26). Further analysis of the
distribution and activities of these pathogens along the colon is needed
to elucidate a mechanism for development of CRC or IBD subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (19).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-\(\alpha\) production than strains from healthy individuals
(10), suggesting the bacteria may increase inflammation in the gut as
well (27). \emph{F. varium} isolated from UC patients caused colonic
ulcers in an experimental mouse model (28). \emph{F. varium} was only
detected in three subjects and two of those samples were isolated form
the proximal mucosa (Figure S2B). \emph{F. varium} is most commonly
isolated from UC patient biopsies from the ileum or cecum (29),
suggesting this species may exhibit preference for the different
environmental conditions of these gastrointestinal sites. Further work
will assess how gut environment may select for species which may then
cause localized disease.

Specific comparisons of our findings to previously published studies of
spatial variation are confounded by the use of bowel preparation
methods. A rare report of a matched-colonoscopy study sampled 18
patient's colonic mucosa and luminal contents prior to and after bowel
cleansing (30). This study found that mucosal and luminal samples were
distinguishable prior to bowel cleansing, but that bowel preparation
resulted in an increase in shared OTUs between each site (30). After
seven days, bowel cleansing not only made the samples more difficult to
distinguish, but it also decreased the diversity observed across sites.
Bowel preparation clearly biases the representation of microbiota
recovered from sampling the lumen or mucosa.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (31). Sequences were curated
using the mothur software as described previously (32). The sequences
were assigned a taxonomic classification using a naive Bayesian
classifier trained using a 16S rRNA gene training set from the Ribosomal
Database Project (RDP) (33) and clustered into operational taxonomic
units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
of a mock community revealed the error rate to be 0.018\%. Samples were
rarefied to 4231 sequences per sample in order to reduce uneven sampling
bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)YC calculator metrics in mothur version 1.39.5 (32).
\(\theta\)YC distances were calculated to determine the dissimilarity
between two samples. Random Forest classification models were built
using the AUCRF R package using a leave-one-subject out approach (34).
For each model the data was split into a 19-subject training set and a
1-subject test set. The model was built and cross-validated using
AUCRFcv on the training set. The model was then tested on the left-out
patient. This process was repeated iteratively for all subjects and
results plotted as Reciever Operator Characteristic curves using the
pROC R package (35). Resultant models were used to identify the OTUs
that were most important for classifying each location. Species-level
information for sequences of interest was obtained by aligning the
sequences to the GenBank nucleotide databse using blastn. The species
name was only used if the identity score was \(\ge\) 99\% over the
full-length of the contig and matched a single reference.

@@

@@

Taxa specific to the distal and proximal mucosa and lumen. The five OTUs
that were most important differentiating the distal and proximal mucosa
(A) and the distal and proximal lumen (B). The vertical lines represent
the median values for each OTU. \newpage

\paragraph{Figure S2}\label{figure-s2}

Location and relative abundance of cancer-associated OTUs. Relative
abundance was calculated and plotted by sample site for each OTU of
interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}

\subsubsection*{References}\label{references}
\addcontentsline{toc}{subsubsection}{References}

@@

\hypertarget{ref-LloydPrice2016}{}
14. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
microbiome. Genome Medicine {[}Internet{]}. Springer Nature; 2016;8.
Available from: \url{https://doi.org/10.1186/s13073-016-0307-y}

\hypertarget{ref-Eckburg2005}{}
15. Eckburg PB. Diversity of the human intestinal microbial flora.
Science {[}Internet{]}. American Association for the Advancement of
Science (AAAS); 2005;308:1635--8. Available from:
\url{https://doi.org/10.1126/science.1110591}

\hypertarget{ref-deCarcer2010}{}
16. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
al. Numerical ecology validates a biogeographical distribution and
gender-based effect on mucosa-associated bacteria along the human colon.
The ISME Journal {[}Internet{]}. Springer Nature; 2010;5:801--9.
Available from: \url{https://doi.org/10.1038/ismej.2010.177}

\hypertarget{ref-Zhang2013}{}
17. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
heterogeneity and co-occurrence patterns of human mucosal-associated
intestinal microbiota. The ISME Journal {[}Internet{]}. Springer Nature;
2013;8:881--93. Available from:
\url{https://doi.org/10.1038/ismej.2013.185}

\hypertarget{ref-Castellarin2011}{}
18. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
Strauss J, et al. Fusobacterium nucleatum infection is prevalent in
human colorectal carcinoma. Genome Research {[}Internet{]}. Cold Spring
Harbor Laboratory Press; 2011;22:299--306. Available from:
\url{https://doi.org/10.1101/gr.126516.111}

\hypertarget{ref-Lee2016}{}
19. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
recovered from colonic biopsies of inflammatory bowel disease patients
in korea. Annals of Laboratory Medicine {[}Internet{]}. Korean Society
for Laboratory Medicine (KAMJE); 2016;36:387. Available from:
\url{https://doi.org/10.3343/alm.2016.36.4.387}

\hypertarget{ref-Hong2011}{}
20. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
Pyrosequencing-based analysis of the mucosal microbiota in healthy
individuals reveals ubiquitous bacterial groups and micro-heterogeneity.
Ahmed N, editor. PLoS ONE {[}Internet{]}. Public Library of Science
(PLoS); 2011;6:e25042. Available from:
\url{https://doi.org/10.1371/journal.pone.0025042}

\hypertarget{ref-Stearns2011}{}
21. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
tract. Scientific Reports {[}Internet{]}. Springer Nature; 2011;1.
Available from: \url{https://doi.org/10.1038/srep00170}

\hypertarget{ref-Sears2014}{}
22. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
from: \url{https://doi.org/10.1016/j.chom.2014.02.007}

\hypertarget{ref-Mima2016}{}
23. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
Fusobacterium nucleatum in colorectal carcinoma tissue according to
tumor location. Clinical and Translational Gastroenterology
{[}Internet{]}. Springer Nature; 2016;7:e200. Available from:
\url{https://doi.org/10.1038/ctg.2016.53}

\hypertarget{ref-Tahara_2014}{}
24. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
al. Fusobacterium in colonic flora and molecular features of colorectal
carcinoma. Cancer Research {[}Internet{]}. American Association for
Cancer Research (AACR); 2014;74:1311--8. Available from:
\url{https://doi.org/10.1158\%2F0008-5472.can-13-1865}

>>>>

2. Department of Family and Community Medicine, College of Medicine,
Pennsylvania State University, Hershey, Pennsylvania 17033

@@

Grant support: Rose and Lawrence C. Page Foundation

Disclosures: nothing to disclose

Word count:

Fig.s:

Supplementary Fig.s:

@@

The microbiome has been implicated in the development of colorectal
cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
of these diseases vary along the axis of the digestive tract. Further,
variation in the structure of the gut microbiota has been associated
with both diseases. Here we profiled the microbiota of the healthy
proximal and distal mucosa and lumen to better understand how bacterial
populations vary along the colon. We used a two-colonoscope approach to
sample proximal and distal mucosal and luminal contents from the colons
of 20 healthy subjects that had not undergone any bowel preparation
procedure. The biopsies and home-collected stool were subjected to 16S
rRNA gene sequencing and Random Forest classification models were built
using taxa abundance and location to identify microbiota specific to
each site. The right mucosa and lumen had the most similar community
structures of the five sites we considered from each subject. The distal
mucosa had higher relative abundance of Finegoldia, Murdochiella,
Peptoniphilus, Porphyromonas and Anaerococcus. The proximal mucosa had
more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas. The
classification model performed well when classifying mucosal samples
into proximal or distal sides (AUC=0.850). Separating proximal and
distal luminal samples proved more challenging (AUC=0.580) and specific
microbiota that differentiated the two were hard to identify. By
sampling the unprepped colon, we identified distinct bacterial
populations native to the proximal and distal sides. Further
investigation of these bacteria may elucidate if and how these groups
contribute to different disease processes on their respective sides of
the colon.

@@

Keywords: microbiome; colon cancer; proximal and distal colon

@@

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (14). Sequences were curated
using the mothur software as described previously (15). The sequences
were assigned a taxonomic classification using a naive Bayesian
classifier trained using a 16S rRNA gene training set from the Ribosomal
Database Project (RDP) (16) and clustered into operational taxonomic
units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
of a mock community revealed the error rate to be 0.018\%. Samples were
rarefied to 4231 sequences per sample in order to reduce uneven sampling
bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)YC calculator metrics in mothur version 1.39.5 (15).
\(\theta\)YC distances were calculated to determine the dissimilarity
between two samples. Random Forest classification models were built
using the AUCRF R package using a leave-one-subject out approach (17).
For each model the data was split into a 19-subject training set and a
1-subject test set. The model was built and cross-validated using
AUCRFcv on the training set. The model was then tested on the left-out
patient. This process was repeated iteratively for all subjects and
results plotted as Reciever Operator Characteristic curves using the
pROC R package (18). Resultant models were used to identify the OTUs
that were most important for classifying each location. Species-level
information for sequences of interest was obtained by aligning the
sequences to the GenBank nucleotide databse using blastn. The species
name was only used if the identity score was \(\ge\) 99\% over the
full-length of the contig and matched a single reference.

@@

Luminal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Fig. 1). Subjects also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on DNA extracted from each
sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (19). Samples had
varying levels of diversity at each site, irrespective of the individual
(Fig. 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

To compare similarity between the proximal and distal sides and within
the lumen and mucosa, we compared the community structure of these sites
based on the relative abundances of individual Operational Taxonomic
Units (OTUs). Across all subjects we observed wide variation when
comparing sample locations (Fig. 3A). Those ranges did not follow a
clear pattern on an individual basis. However, when comparing median
dissimilarity between the communities found in the proximal lumen and
mucosa, the proximal and distal lumen, the proximal and distal mucosa,
and the distal lumen and mucosa, we found that the proximal lumen and
mucosa were most similar to each other than to the other samples (P
\textless{} 0.005, Wilcoxon, BH adjustment).

Next, we compared the luminal and mucosal samples to the fecal sample of
each subject. Amidst the large inter-subject variation, we did identify
significantly less dissimilarity between the distal luminal sample and
the feces (Fig. 3B, P \textless{} 0.05, Wilcoxon, BH adjustment).
Furthermore, there was an even larger difference in the communities
found in the distal mucosa compared to the fecal communities, indicating
that the mucosa is as different from the stool as compared to lumen (P
\textless{} 0.0005, Wilcoxon, BH adjustment). These results suggest that
the contents of the distal lumen were most representative of the
subjects' feces, and the mucosal microbiota are distinct from the fecal
and luminal communities.

To determine what factors may have driven the differences seen among the
samples, we compared the community dissimilarity between samples from
all subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to matched samples from the other subjects
(Fig. 3C); this is consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (20--22). Thus
interpersonal variation drove the differences between samples more than
whether the sample came from the proximal or distal side of the colon or
from the lumen or mucosa.

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We used
10-fold cross validation to build the first model to classify the
luminal versus mucosal samples for the proximal and distal sides,
independently (Fig. 4A). The models performed well when classifying
these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
were most predictive of each site were identified by their greatest mean
decrease in accuracy when removed from the model. For distinguishing the
proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
included in the best model (Fig. 5A). The model to differentiate between
the distal lumen and mucosa included OTUs affiliated with the
\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
\emph{Anaerococcus} (Fig. 5B). These results indicated that there were
fine differences between the different sites of the colon, and that
these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples using 10-fold cross validation. The model
performed best when distinguishing the proximal versus distal mucosa
(Fig. 4B, AUC = 0.850) whereas the model to differentiate between the
proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
included in the model differentiating the distal and proximal mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Fig.
6A). The model that attempted to separate the the proximal and distal
lumen included OTUs affiliated with the \emph{Bacteroides},
\emph{Clostridium IV} and \emph{Oscillibacter} (Fig. 6B). Interestingly,
\emph{Anaerococcus} and \emph{Finegoldia} were distinct between the
mucosa and lumen and also helped to differentiate between the proximal
and distal sides.

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors
(23). There were 14 samples from 8 subjects with the \emph{F. nucleatum
subsp. animalis} sequences. Of the samples with the highest relative
abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
were from the proximal mucosa and three from the distal mucosa
(Supplementary Fig. S1A). The second most frequent \emph{Fusobacterium}
sequence was OTU472, which aligned with 99\% identity to \emph{F.
varium}. In addition to \emph{F. nucleatum}, \emph{F. varium} has been
associated with IBD (24). Four subjects harbored \emph{F. varium} and
the samples were split evenly between the proximal and distal mucosa
(Supplementary Fig. S1B). OTU152 was similar to the members of the
\emph{Porphyromonas} genus and the most frequent sequence in that OTU
aligned to \emph{Porphyromonas asacharolytica} (99\% over full length),
another bacterium commonly detected and isolated from colorectal tumors.
OTU152 was only detected on the distal mucosa, and in fact was one of
the OTUs the classification model identified as separating distal and
proximal sides (Supplementary Fig. S1C). Among the 11 distal mucosa
samples that were positive for \emph{P. asacharolytica}, the relative
abundances for this OTU ranged from 0.01\% to 16\%. Thus,
disease-associated OTUs could be found in our sample set of 20 healthy
individuals.

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, casued by differences in age, sex, diet, but also
spatial ``patchiness'' in the gut microbiome. One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(25). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (26). That
said, across our samples, the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to luminal content in humans (4). Hence, the conclusions
we were able to draw from phyla analysis may have been impacted by
inter-subject patchiness.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (27). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} along the
colon, but given its anaerobic lifestyle and asacharolytic metabolism,
it is not surprising that it resides in the less-oygen-rich and
protein-rich distal mucosa (4). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreased along the colon to rectum (28). In another study,
\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
mutations, molecular features of proximal CRC (29). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in the present study, the
bacterium was spread across the proximal mucosa, distal lumen and distal
mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
suggests that \emph{Fusobacteria} species are more commonly found on
proximal tumors and in biofilms, indicating that it is not only the
presence of the bacterium but the structure of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
only co-occur on CRC tumors but also to synergize to promote
tumorigenesis in an oral cancer model (30) (31). Further analysis of the
distribution and activities of these pathogens along the colon is needed
to elucidate a mechanism for development of CRC or IBD subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (24).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-\(\alpha\) production than strains from healthy individuals
(10), suggesting the bacteria may increase inflammation in the gut as
well (32). \emph{F. varium} isolated from UC patients caused colonic
ulcers in an experimental mouse model (33). \emph{F. varium} was only
detected in three subjects and two of those samples were isolated form
the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
commonly isolated from UC patient biopsies from the ileum or cecum (34),
suggesting this species may exhibit preference for the different
environmental conditions of these gastrointestinal sites. Further work
will assess how gut environment may select for species which may then
cause localized disease.

Specific comparisons of our findings to previously published studies of
spatial variation are confounded by the use of bowel preparation
methods. A rare report of a matched-colonoscopy study sampled 18
patient's colonic mucosa and luminal contents prior to and after bowel
cleansing (35). This study found that mucosal and luminal samples were
distinguishable prior to bowel cleansing, but that bowel preparation
resulted in an increase in shared OTUs between each site (35). After
seven days, bowel cleansing not only made the samples more difficult to
distinguish, but it also decreased the diversity observed across sites.
Bowel preparation clearly biases the representation of microbiota
recovered from sampling the lumen or mucosa.

@@

\subsubsection{References}\label{references}

@@

\hypertarget{ref-Kozich2013}{}
14. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq illumina sequencing
platform. Applied and Environmental Microbiology {[}Internet{]}.
American Society for Microbiology; 2013;79:5112--20. Available from:
\url{https://doi.org/10.1128/aem.01043-13}

\hypertarget{ref-Schloss2009}{}
15. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology {[}Internet{]}.
American Society for Microbiology; 2009;75:7537--41. Available from:
\url{https://doi.org/10.1128/aem.01541-09}

\hypertarget{ref-Wang2007}{}
16. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive bayesian classifier
for rapid assignment of rRNA sequences into the new bacterial taxonomy.
Applied and Environmental Microbiology {[}Internet{]}. American Society
for Microbiology; 2007;73:5261--7. Available from:
\url{https://doi.org/10.1128/aem.00062-07}

\hypertarget{ref-Calle_2011}{}
17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
for genomic profiling with random forest. Human Heredity {[}Internet{]}.
S. Karger AG; 2011;72:121--32. Available from:
\url{https://doi.org/10.1159\%2F000330778}

\hypertarget{ref-Robin_2011}{}
18. Robin X, Turck N, Hainard A, Tiberti N, Lisacek F, Sanchez J-C, et
al. pROC: An open-source package for r and s to analyze and compare ROC
curves. BMC Bioinformatics {[}Internet{]}. Springer Nature; 2011;12:77.
Available from: \url{https://doi.org/10.1186\%2F1471-2105-12-77}

\hypertarget{ref-LloydPrice2016}{}
19. Lloyd-Price J, Abu-Ali G, Huttenhower C. The healthy human
microbiome. Genome Medicine {[}Internet{]}. Springer Nature; 2016;8.
Available from: \url{https://doi.org/10.1186/s13073-016-0307-y}

\hypertarget{ref-Eckburg2005}{}
20. Eckburg PB. Diversity of the human intestinal microbial flora.
Science {[}Internet{]}. American Association for the Advancement of
Science (AAAS); 2005;308:1635--8. Available from:
\url{https://doi.org/10.1126/science.1110591}

\hypertarget{ref-deCarcer2010}{}
21. Cárcer DA de, Cuív PÓ, Wang T, Kang S, Worthley D, Whitehall V, et
al. Numerical ecology validates a biogeographical distribution and
gender-based effect on mucosa-associated bacteria along the human colon.
The ISME Journal {[}Internet{]}. Springer Nature; 2010;5:801--9.
Available from: \url{https://doi.org/10.1038/ismej.2010.177}

\hypertarget{ref-Zhang2013}{}
22. Zhang Z, Geng J, Tang X, Fan H, Xu J, Wen X, et al. Spatial
heterogeneity and co-occurrence patterns of human mucosal-associated
intestinal microbiota. The ISME Journal {[}Internet{]}. Springer Nature;
2013;8:881--93. Available from:
\url{https://doi.org/10.1038/ismej.2013.185}

\hypertarget{ref-Castellarin2011}{}
23. Castellarin M, Warren RL, Freeman JD, Dreolini L, Krzywinski M,
Strauss J, et al. Fusobacterium nucleatum infection is prevalent in
human colorectal carcinoma. Genome Research {[}Internet{]}. Cold Spring
Harbor Laboratory Press; 2011;22:299--306. Available from:
\url{https://doi.org/10.1101/gr.126516.111}

\hypertarget{ref-Lee2016}{}
24. Lee Y, Eun CS, Lee AR, Park CH, Han DS. FusobacteriumIsolates
recovered from colonic biopsies of inflammatory bowel disease patients
in korea. Annals of Laboratory Medicine {[}Internet{]}. Korean Society
for Laboratory Medicine (KAMJE); 2016;36:387. Available from:
\url{https://doi.org/10.3343/alm.2016.36.4.387}
